---
title: "Missing Data Simulation Study"
output: html_document
date: "2026-06-09"
---
#Packages Used
```{r}
library(mice)
```

#Build a simulated RCBD Dataset
```{r}
#Set Seed
set.seed(42)

#Set initial parameters (these can change)
n_treatments <- 8
n_blocks <- 7

#Define Treatment and Block vectors
treatments <- paste0("T", 1:n_treatments)
blocks     <- paste0("Block", 1:n_blocks)

#Treatment and block effects 
treatment_effects <- c(0,0,0,0,0,0,0)
block_effects <- c(1,1,1,1,1,1,1)
error <- rnorm(n_treatments * n_blocks, 0, 3)

#Build Dataframe
RCDB_df <- expand.grid(
  Treatment = treatments,
  Block     = blocks)

#Treatment Effects  
RCDB_df$treatment_effect <- treatment_effects[as.numeric(factor(RCDB_df$Treatment))]

#Block Effects
RCDB_df$block_effect <- block_effects[as.numeric(factor(RCDB_df$Block))]

#Response 
RCDB_df$response <- RCDB_df$treatment_effect + 
                    RCDB_df$block_effect + 
                    error
#Clean Up
RCDB_df$treatment_effect <- NULL
RCDB_df$block_effect     <- NULL

RCDB_df$Treatment <- factor(RCDB_df$Treatment)
RCDB_df$Block     <- factor(RCDB_df$Block)

#Preview
head(RCDB_df)

```

#MCAR RCBD data
```{r}
#Create new df to manipulate 
MCAR_RCBD_df <- RCDB_df

#Define percentage of missing data
missing_percentage <- 0.10

#Apply missing percentage to new df 
n_missing <- round(nrow(MCAR_RCBD_df) * missing_percentage)

missing_rows <- sample(1:nrow(MCAR_RCBD_df), size = n_missing, replace = FALSE)

MCAR_RCBD_df$response[missing_rows] <- NA

head(MCAR_RCBD_df)

```

#Mean Imputation
```{r}
mean_MCAR <- mean(MCAR_RCBD_df$response, na.rm = TRUE)

mean_MCAR_RCBD_df <- MCAR_RCBD_df

mean_MCAR_RCBD_df$response[is.na(mean_MCAR_RCBD_df$response)] <- mean_MCAR
```

#Multiple Imputation Using MICE
```{r}
imputed <- mice(MCAR_RCBD_df, m = 5, method = 'pmm', seed = 42)

imputed_MCAR <- completed_df <- complete(imputed, action = 1)

completed_all <- complete(imputed, action = "long")
```

#All of these codes are used to compare a single run, whereas we want to look at several iterations
#Therefore new code must be needed in order to look at it over a certain number for simulations. 


#Build function which builds the 3 RCBDs we are looking at
```{r}
##################################################
# Packages
##################################################

library(mice)

##################################################
# Simulation Function
##################################################
n_treatments <- 8
    n_blocks <- 7
    block_effects <- rnorm(n_blocks, mean = 0, sd = 3)
    treatment_effects <- c(0,0,0,0,0,0,0,0)

one_simulation <- function(
    treatment_effects,
    block_effects = rnorm(n_blocks, mean = 0, sd = 3),
    missing_pct = 0.10,
    sigma = 3){

  ################################################
  # Generate RCBD
  ################################################

  df <- expand.grid(
    Treatment = factor(paste0("T",1:n_treatments)),
    Block = factor(paste0("B",1:n_blocks))
  )


  df$response <-
    treatment_effects[as.numeric(df$Treatment)] +
    block_effects[as.numeric(df$Block)] +
    rnorm(nrow(df),0,sigma)

  ################################################
  # Complete Data Analysis
  ################################################

  fit_complete <- aov(
    response ~ Treatment + Block,
    data = df
  )

  p_complete <-
    anova(fit_complete)$`Pr(>F)`[1]

  ################################################
  # MCAR Missingness
  ################################################

  missing_df <- df

  n_missing <- round(
    nrow(missing_df) * missing_pct
  )

  missing_rows <- sample(
    seq_len(nrow(missing_df)),
    n_missing,
    replace = FALSE
  )

  missing_df$response[missing_rows] <- NA

  ################################################
  # Mean Imputation
  ################################################

  mean_df <- missing_df

  mean_df$response[
    is.na(mean_df$response)
  ] <- mean(
    mean_df$response,
    na.rm = TRUE
  )

  fit_mean <- aov(
    response ~ Treatment + Block,
    data = mean_df
  )

  p_mean <-
    anova(fit_mean)$`Pr(>F)`[1]
  
  ################################################
  # LINEAR REGRESSION IMPUTATION
  ################################################
  
  linear_df <- missing_df
  
  linear_df$response[is.na(missing_df$response)] <- predict(
  lm(response ~ Treatment + Block, data = linear_df),
  newdata = missing_df[is.na(linear_df$response), ]
)
  
 fit_linear <-  aov(
    response ~ Treatment + Block,
    data = linear_df)
 
 p_linear <- anova(fit_linear)$`Pr(>F)`[1]
  
  
  ################################################
  # MICE Imputation
  ################################################

  imp <- mice(
    missing_df,
    m = 5,
    method = c("", "", "pmm"),
    printFlag = FALSE
  )

  mice_df <- complete(imp,1)

  fit_mice <- aov(
    response ~ Treatment + Block,
    data = mice_df
  )

  p_mice <-
    anova(fit_mice)$`Pr(>F)`[1]

  ################################################
  # Return p-values
  ################################################

 return( p_val <- c(
    complete = p_complete,
    mean_imp = p_mean,
    mice = p_mice,
    linear_imp = p_linear
  )
 )
  
}
one_simulation(treatment_effects, block_effects)
```


#Replicate the function created in the chunk above
```{r}
set.seed(123)
n_sims <- 5000
alpha  <- 0.05
 
# Run all simulations and collect p-values into a matrix (n_sims x 3)
sim_results <- t(replicate(
  n_sims,
  one_simulation(treatment_effects,block_effects)
))
 
# Compute Type I error rate = proportion of p-values < alpha under H0
type1_error <- colMeans(sim_results < alpha, na.rm = TRUE)
 
# Summary table
type1_table <- data.frame(
  Method        = c("Complete Data", "Mean Imputation", "MICE (Pooled)", "Linear Imputation"),
  Type_I_Error  = round(type1_error, 4),
  row.names     = NULL
)
 
cat("\n==== Type I Error Rate Comparison (alpha =", alpha, ", n_sims =", n_sims, ") ====\n")
print(type1_table)
```



