---
title: "RCBD Outlier Simulation Study"
output: html_document
---

# Packages Used
```{r}
library(tidyverse)
library(broom)
```

# Mixture Model: Contaminated Normal Outliers
```{r}
outlier_rcbd1 <- function(blocks, treatments, sigma_e, sigma_b,
                         sigma_G, pi) {

  data <- expand.grid(
    block = factor(1:blocks),
    treatment = factor(1:treatments)
  )

  # Simulation setup under the Null Hypothesis (no true treatment effects)
  alpha_t <- rep(0, treatments)
  beta_b <- rnorm(blocks, 0, sigma_b)

  n <- nrow(data)

  # Generate binary contamination indicator using a uniform distribution
  outlier <- runif(n) < pi

  # Normal observations drawn from N(0, sigma_e^2); outliers drawn from N(0, sigma_G^2)
  e <- rnorm(n, 0, sigma_e)
  e[outlier] <- rnorm(sum(outlier), 0, sigma_G)

  data$response <-
    alpha_t[as.integer(data$treatment)] +
    beta_b[as.integer(data$block)] +
    e

  # Additive Two-Way ANOVA Evaluation
  model <- aov(response ~ treatment + block, data = data)
  anova_tbl <- anova(model) |> tidy()

  F_stat <- anova_tbl |>
    filter(term == "treatment") |>
    pull(statistic)
  
  p_value <- anova_tbl |>
    filter(term == "treatment") |>
    pull(p.value)

  # Monitor operational Type I Error rate limits (alpha = 0.05)
  reject_F <- p_value < 0.05

  # Track Family-Wise Error Rate (FWER) using adjusted Tukey's p-values
  tukey_flag <- TukeyHSD(model, "treatment")$treatment |>
    as.data.frame() |>
    pull(`p adj`) |>
    (\(x) any(x < 0.05))()

  R2 <- summary(lm(response ~ treatment + block, data))$r.squared

  # Extract structural mean squares to determine diagnostic variance inflation
  MS_error <- anova_tbl |>
    filter(term == "Residuals") |>
    pull(meansq)
  
  MS_treatment <- anova_tbl |>
    filter(term == "treatment") |>
    pull(meansq)

  tibble(
    F_stat = F_stat,
    reject_F = reject_F,
    R2 = R2,
    tukey_reject = tukey_flag,
    MSE = MS_error,
    MSa = MS_treatment,
  )
}
```

# EXPERIMENT 1: CONTAMINATED NORMAL OUTLIERS (EXTREME VS MILD SCALE)
```{r}
set.seed(100)

# Establish vector grid for contamination proportion bounds (0% to 10%)
pi_grid <- seq(0, 0.1, by = 0.01)

# Scenario A: Contamination using extreme normal outliers (sigma_G = 100)
outlier_pi_curve <- map_dfr(pi_grid, function(pi) {
  map_dfr(1:10000, ~ outlier_rcbd1(7, 8, 3, 3, 100, pi)) |>
    summarise(
      pi = pi,
      type1_error = mean(reject_F),
      mean_F = mean(F_stat),
      var_F = var(F_stat),
      mean_MSa = mean(MSa),
      var_MSa = var(MSa),
      mean_MSE = mean(MSE),
      var_MSE = var(MSE),
      mean_R2 = mean(R2),
      tukey_fwer = mean(tukey_reject)
    )
})
```


# Display Type I Error robustness behavior across extreme variance scales
```{r}
ggplot(outlier_pi_curve,
       aes(x = pi, y = type1_error)) +
  geom_line() +
  geom_point() +
  labs(
    x = expression(pi),
    y = "Type I Error Rate",
    title = "Type I Error Rate for Data With Extreme Normal Outliers"
  ) +
  ylim(0, 0.1) +
  theme_bw()
```


# Mixture Model: Deterministic (Constant) Outliers
```{r}
outlier_rcbd2 <- function(blocks, treatments, sigma_e, sigma_b, delta_G, pi) {

  data <- expand.grid(
    block = factor(1:blocks),
    treatment = factor(1:treatments)
  )

  alpha_t <- rep(0, treatments)
  beta_b <- rnorm(blocks, 0, sigma_b)

  n <- nrow(data)
  outlier <- runif(n) < pi

  # Contaminated elements receive an exact constant magnitude offset (delta_G)
  e <- rnorm(n, 0, sigma_e)
  e[outlier] <- rep(delta_G, sum(outlier))

  data$response <-
    alpha_t[as.integer(data$treatment)] +
    beta_b[as.integer(data$block)] +
    e

  model <- aov(response ~ treatment + block, data = data)
  anova_tbl <- anova(model) |> tidy()

  F_stat <- anova_tbl |>
    filter(term == "treatment") |>
    pull(statistic)
  
  p_value <- anova_tbl |>
    filter(term == "treatment") |>
    pull(p.value)

  reject_F <- p_value < 0.05

  tukey_flag <- TukeyHSD(model, "treatment")$treatment |>
    as.data.frame() |>
    pull(`p adj`) |>
    (\(x) any(x < 0.05))()

  R2 <- summary(lm(response ~ treatment + block, data))$r.squared

  MS_error <- anova_tbl |>
    filter(term == "Residuals") |>
    pull(meansq)
  
  MS_treatment <- anova_tbl |>
    filter(term == "treatment") |>
    pull(meansq)

  tibble(
    F_stat = F_stat,
    reject_F = reject_F,
    R2 = R2,
    tukey_reject = tukey_flag,
    MSE = MS_error,
    MSa = MS_treatment,
  )
}
```


# EXPERIMENT 2: CONSTANT OUTLIER SIMULATIONS
```{r}
set.seed(100)

outlier_pi_curve2 <- map_dfr(pi_grid, function(pi) {
  map_dfr(1:10000, ~ outlier_rcbd2(7, 8, 3, 3, 100, pi)) |>
    summarise(
      pi = pi,
      type1_error = mean(reject_F),
      mean_F = mean(F_stat),
      var_F = var(F_stat),
      mean_MSa = mean(MSa),
      var_MSa = var(MSa),
      mean_MSE = mean(MSE),
      var_MSE = var(MSE),
      mean_R2 = mean(R2),
      tukey_fwer = mean(tukey_reject)
    )
})
```


# Display Type I Error performance curves under constant directional asymmetry
```{r}
ggplot(outlier_pi_curve2,
       aes(x = pi, y = type1_error)) +
  geom_line() +
  geom_point() +
  labs(
    x = expression(pi),
    y = "Type I Error Rate",
    title = "Type I Error Rate for Data With Extreme Constant Outliers"
  ) +
  ylim(0, 0.1) +
  theme_bw()
```


# Mixture Model: Heavy-Tailed Cauchy Outliers
```{r}
outlier_rcbd3 <- function(blocks, treatments, sigma_e, sigma_b,
                         gamma_G, pi) {

  data <- expand.grid(
    block = factor(1:blocks),
    treatment = factor(1:treatments)
  )

  alpha_t <- rep(0, treatments)
  beta_b <- rnorm(blocks, 0, sigma_b)

  n <- nrow(data)
  outlier <- runif(n) < pi

  e <- rnorm(n, 0, sigma_e)

  # Contaminated profiles modeled using Cauchy distribution scaling indexes
  e[outlier] <- rcauchy(sum(outlier), location = 0, scale = gamma_G)

  data$response <-
    alpha_t[as.integer(data$treatment)] +
    beta_b[as.integer(data$block)] +
    e

  model <- aov(response ~ treatment + block, data = data)
  anova_tbl <- anova(model) |> tidy()

  F_stat <- anova_tbl |>
    filter(term == "treatment") |>
    pull(statistic)
  
  p_value <- anova_tbl |>
    filter(term == "treatment") |>
    pull(p.value)

  reject_F <- p_value < 0.05

  tukey_flag <- TukeyHSD(model, "treatment")$treatment |>
    as.data.frame() |>
    pull(`p adj`) |>
    (\(x) any(x < 0.05))()

  R2 <- summary(lm(response ~ treatment + block, data))$r.squared

  MS_error <- anova_tbl |>
    filter(term == "Residuals") |>
    pull(meansq)
  
  MS_treatment <- anova_tbl |>
    filter(term == "treatment") |>
    pull(meansq)

  tibble(
    F_stat = F_stat,
    reject_F = reject_F,
    R2 = R2,
    tukey_reject = tukey_flag,
    MSE = MS_error,
    MSa = MS_treatment,
  )
}
```

# EXPERIMENT 3: HEAVY-TAILED CAUCHY OUTLIER SIMULATIONS
```{r}
set.seed(100)

outlier_pi_curve3 <- map_dfr(pi_grid, function(pi) {
  map_dfr(1:10000, ~ outlier_rcbd3(7, 8, 3, 3, 25, pi)) |>
    summarise(
      pi = pi,
      type1_error = mean(reject_F),
      mean_F = mean(F_stat),
      var_F = var(F_stat),
      mean_MSa = mean(MSa),
      var_MSa = var(MSa),
      mean_MSE = mean(MSE),
      var_MSE = var(MSE),
      mean_R2 = mean(R2),
      tukey_fwer = mean(tukey_reject)
    )
})
```


# Display Type I Error performance curves under heavy-tailed (Cauchy) outliers
```{r}
ggplot(outlier_pi_curve3,
       aes(x = pi, y = type1_error)) +
  geom_line() +
  geom_point() +
  labs(
    x = expression(pi),
    y = "Type I Error Rate",
    title = "Type I Error Rate for Data With Extreme Cauchy Outliers"
  ) +
  ylim(0, 0.1) +
  theme_bw()
```

